A primer to numerical simulations: The perihelion motion of Mercury

### 2018-04-18 tex/MercuryNumerical.tex
Add my mustard to the text

The dynamics is controlled by Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory by solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be fixed --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit1}
		Different Mercury orbits for the purely Newtonian gravity and
		 $\Delta t = \Delta t_0  / 20$ for (a) and $\Delta t = \Delta t_0  \times 2$ for (b), 
		 where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps;
		Here, the reference time step is given by $\Delta t_0$ defined via Eq.~(\ref{eq:check}) as explained
		in the text.
		The images are screenshots of the simulation that is described below (with modified colors).
	}
\end{figure}

This procedure can only work if $\Delta t$ is sufficiently small.
One way to estimate whether $\Delta t$ is small enough is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that was neglected reads $(1/2)a(t){(\Delta t)}^2$.
The impact of the size of $\Delta t$ on the simulation is illustrated in Fig.~\ref{fig:MercuryOrbit1} where
the force employed in Newtonian gravity as detailed in the next section. While for the
calculations for the left panel $\Delta t$
is chosen significantly below $\Delta t_0=2m |\vec v(0)| /   |\vec F(\vec r(0)) |$ ($cf.$ Eq.~(\ref{eq:check})), for the
right panel it was chosen twice as large. Accordingly only the trajectory in the left channel shows the characteristic
feature of a $1/r^2$--force of an ellipse fixed in space --- the trajectory in the right channel does not reproduce itself in successive revolutions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
In the following, we present code examples which tested for \python{} version \texttt{3.6.4} and \vpython{} version \texttt{7.3.2}.
We provide online instructions for setting up \python{} and \vpython{} on different operating systems \cite{scripts}.
In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.

@@

In Section~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution \cite{scripts}.
These instructions can be cut down or left out depending on available time and numerical/computational versatility of the students.
This could be achieved by the following changes or additions to the concept presented in Section~\ref{sec:Numerical Implementation}.

@@

The idea that the numerical study of the perihelion motion of Mercury is an interesting subject for high school students was taken from a project work of one of the authors (JH).
The project was presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
Based on this study, the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} and used in the Sch\"ulerakademie Teilchenphysik in 2015, which aims at high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
This academy is financed by the German Research Society (DFG) as the outreach branch of a research grant focussing on basic research in particle and nuclear physics.
One of the authors (CH) is acting as a Principal Investigator for the outreach activities within this scheme.
The Sch\"ulerakademie Teilchenphysik runs biannually at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
The academy comprises lectures, a tour to the particle accelerator COSY and the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
Given the setting the participating students are highly motivated and eager to understand.

The course presented in this paper is offered as one out of three hands-on project options and was prepared by a lecture to the whole group explaining the basics of numerical simulations.
Typically about one half of the participants chose to work on the numerical project --- out of those about one half had previous programming experience.
Most of the participants had already learned about derivatives and vector calculus in school.
Given the positive to enthusiastic feedback we have received from the participants on the numerical course, we decided to make the course a permanent part of the academy.
It was used in the 2017 with the same success.

The numerical course was allocated for one work day (seven hours) and the students worked in pairs.
Each group worked on a laptop provided by us, where the necessary software was preinstalled.
Four of the authors (IH, CK, CM, JLW) acted as supervisors for the students. 
The course was initiated with a brief summary of previously presented methods, motivations and objectives of the simulation.
In the first two hours the students were introduced to \python{} and \vpython{}.
In the following part the students worked with an initial code template and were motivated to come up with own solutions to fulfill the objectives.
A large part of the material covered during this day is presented in Sec.~\ref{sec:Numerical Implementation}.
While a majority of the participants where able to visually simulate the motion of mercury including forces from GR, only a few participants started to implement the quantitative extraction of the perihelion motion --- not to speak of uncertainty estimates.
Once the basic problem was tackled some students deviated from the suggested path and for instance studied the problem with one additional planet or different starting parameters.
This is a very interesting problem by itself for there are parameter ranges where the classical three-body system develops chaotic features.
We encouraged such explorations as well.
The reason why we included the additional material in the paper is that we wanted to provide teachers with sufficient material to probably even fill a working group outside the general curriculum running for a longer period.

@@

\begin{figure}[htb]
\lstinputlisting[language=Python,basicstyle=\linespread{1}\footnotesize\ttfamily]{mercury_orbit_base.py}
\end{figure}
Further examples and template files can be found online \cite{scripts}.

>>>>

The dynamics is controlled by Newton's second law
\begin{equation}
\vec F(\vec r, t) = \frac{d}{dt}(m \vec v) \ .
\end{equation}
If the mass does not change~\footnote{%
	A well known example where $m$ does change with time is a rocket, whose mass decreases as fuel is burned.
} with time, this reduces to the well known
\begin{equation}
\vec F(\vec r) = m \vec a(t) = m\dot{\vec v}(t) = m\ddot{\vec{r}}(t) \ . \label{eq:newton2}
\end{equation}
Note that in general the force could depend also on the time or the velocity.
Here, we restrict ourselves to the case relevant for our example, where the force depends on the location only.
Therefore, as soon as the force $\vec F(\vec r)$ is known for all $\vec r$ one can in principle calculate the trajectory by solving Eq.~(\ref{eq:newton2}) for $\vec r(t)$.
Sometimes this requires some advanced knowledge of mathematics and sometimes no closed form solution exists.
However, alternatively one can calculate the whole trajectory of some test body that experiences this force by a successive application of the rules given in Eqs.~(\ref{eq:vdef}) and (\ref{eq:adef}):
\begin{enumerate}
\item For a given time $t$, where $\vec r(t)$ and $\vec v(t)$ are known, use the force to calculate $\vec a(t) = \vec F(\vec r(t)) / m$, see Eq.~(\ref{eq:newton2}).
\item Use Eq.~(\ref{eq:vdef}) to calculate $\vec r(t+\Delta t)$.
\item Use Eq.~(\ref{eq:adef}) to calculate $\vec v(t+\Delta t)$.
\item Go back to (i) with $t\to t+\Delta t$.
\end{enumerate}
Clearly to initiate the procedure at some time $t_0$ both $\vec r(t_0)$ as well as $\vec v(t_0)$ must be fixed --- the  trajectories depend on these initial conditions~\footnote{%
	In general, a differential equation of $n^{\rm th}$ degree (where the highest derivative is order $n$) needs $n$ initial conditions specified.
	For $n=2$ those are often chosen as location and velocity at some starting time, but one may as well pick two locations at different times.%
}.
\begin{figure}[htb]
	\centering
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/a0T5dt20.png}
		\caption{\label{fig:MercuryOrbit-a0-small}}
	\end{subfigure}
	~
	\begin{subfigure}[c]{0.22\textwidth}
		\includegraphics[width=\textwidth]{figs/num-err.png}
		\caption{\label{fig:MercuryOrbit-a0-small-dt-large}}
	\end{subfigure}
	\captionsetup{singlelinecheck=off}
	\caption[]{\label{fig:MercuryOrbit1}
		Different Mercury orbits for the purely Newtonian gravity~(\ref{eq:newton_grav}) and
		 $\Delta t = \Delta t_0  / 20$ for (a) and $\Delta t = \Delta t_0  \times 2$ for (b),
		 where the black line represents the orbit of (a) and the red line is the orbit for the larger time steps.
		The reference time step $\Delta t_0$ is given by Eq.~(\ref{eq:check}) as explained
		in the text.
		The images are screenshots of the simulation that is described below (with modified colors).
	}
\end{figure}

This procedure can only work if $\Delta t$ is sufficiently small.
One way to estimate whether $\Delta t$ is small enough is to verify whether the relation
\begin{equation}
|\vec v(t)| \gg \frac12|\vec a(t)|\Delta t = \frac{1}{2m} |\vec F(\vec r(t)) |\Delta t\
\label{eq:check}
\end{equation}
holds. This follows from Eq.~(\ref{eq:vdef}) where the first term that was neglected reads $(1/2)a(t){(\Delta t)}^2$.
The impact of the size of $\Delta t$ on the simulation is illustrated in Fig.~\ref{fig:MercuryOrbit1} where
only Newtonian gravity was used. While for the
calculations for the left panel $\Delta t$
is chosen significantly below $\Delta t_0=2m |\vec v(0)| /   |\vec F(\vec r(0)) |$ ($cf.$ Eq.~(\ref{eq:check})), for the
right panel it was chosen twice as large. Accordingly, only the trajectory in the left panel shows the characteristic
feature of a $1/r^2$--force of an ellipse fixed in space. The trajectory in the right panel does not reproduce itself in successive revolutions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Example: General Relativity and the perihelion motion of Mercury}\label{sec:gr}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
For this concrete example the starting point for the force is Newton's law of gravitation
\begin{equation}
\vec F_N(\vec r) = - \frac{G_N m M_\odot}{r^2} \frac{\vec r}{r}\ ,\label{eq:newton_grav}
\end{equation}
where $G_N=6.67\times 10^{-11}$ m$^3$kg$^{-1}$s$^{-2}$ is the Newtonian constant of gravitation, $m$ is the mass of Mercury and $M_\odot=1.99\times 10^{30}$ kg is the mass of the Sun.
In addition $r=|\vec r(t)|$ denotes the distance between Sun and Mercury, when we assume that the Sun is infinitely heavier than Mercury and located at the center of the coordinate system.
Although this is not exact, it is a good approximation because $m/M_\odot\sim 10^{-8}$.
For later convenience we introduce the Schwarzschild radius of the Sun
\begin{equation}
r_S=\frac{2G_N  M_\odot}{c^2} = 2.95 \ \mbox{km} \ , \label{rsdef}
\end{equation}
where $c=3.00\times 10^8$ m/s denotes the speed of light.
Note that $r_S$ is the characteristic length scale of the gravitational field of the Sun for --- up to the prefactor --- one can not form another quantity with dimensions of a length from $G_N$, $M_\odot$ and $c$.
The Schwarzschild radius is also an important quantity to characterize black holes; this is however not relevant for the discussion at hand.
With this, Newton's second law reads
\begin{equation}
\ddot{\vec r}      = - \frac{c^2}{2}\left(\frac{r_S}{r^2}\right)\frac{\vec r}{r} \, . \label{eq:newton}
\end{equation}
An attractive force that vanishes at large distances leads in general, depending on the initial conditions, either to bounded or to open orbits.
In the latter case the planet simply disappears from the Sun, since a given force can capture only bodies with small enough momenta.
The students may study those scenarios within their simulations by varying the start velocity while keeping the start location fixed.

@@

In this section we present the numerical implementation as well as the visualization of planetary trajectories and in
particular the perihelion motion of Mercury.
We choose \python{}~\cite{Python} as programming language because \python{} is easy to learn, intuitive to understand and an open source language.
No prior knowledge of \python{} or any other programming language is required, as we explain all necessary steps to create the simulation.
In the following, we present code examples which tested for \python{} version \texttt{3.6.4} and \vpython{} version \texttt{7.3.2}.
We provide online instructions for setting up \python{} and \vpython{} on different operating systems~\cite{scripts}.
In \ref{sec:code}, we also provide a working example as well as possible extensions and template files online.

@@

In Section~\ref{sec:Numerical Implementation} we suggested to present the material by using a template as well as a step by step instruction to guide the students through the problem solution~\cite{scripts}.
These instructions can be cut down or left out depending on available time and numerical/computational versatility of the students.
This could be achieved by the following changes or additions to the concept presented in Section~\ref{sec:Numerical Implementation}.

@@

Work on the subject began with a project work of one of the authors (JH).
The project was presented at the M\"adchen-Gymnasium J\"ulich (a German high school) in 11$^{\rm th}$ grade in 2014.
Based on this study, the course was developed as outlined in Sec.~\ref{sec:Numerical Implementation} and used in the Sch\"ulerakademie Teilchenphysik in 2015, which aims at high school students from 10$^{\rm th}$ to 13$^{\rm th}$ grade.
This academy is financed by the German Research Society (DFG) as the outreach branch of a research grant focussing on basic research in particle and nuclear physics.
One of the authors (CH) is acting as a Principal Investigator for the outreach activities within this scheme.
The Sch\"ulerakademie Teilchenphysik runs biennially at the Science Center Overbach in J\"ulich for four days and has about 25 participants coming from various parts of Germany.
The academy comprises lectures, a tour to the particle accelerator COSY and the high performance computing facility at the Research Center J\"ulich as well as one full day of practical work.
Given the setting, the participating students are highly motivated and eager to understand.

The course presented in this paper is offered as one out of three hands-on project options.
A lecture to the whole group explaining the basics of numerical simulations laid down some foundations in addition to what is presented in this paper.
Given the positive to enthusiastic feedback we have received from the participants on the course after its first installation in 2015, we decided to make the course a permanent part of the academy.
In 2017 we received similarly positive feedback.
About one half of the participants of each year chose to work on the numerical project --- out of those about one half had previous programming experience.
Most of the participants had already learned about derivatives and vector calculus in school.

The numerical course was allocated for one work day (seven hours) and the students worked in pairs.
Each group worked on a laptop provided by us, where the necessary software was pre-installed.
Four of the authors (IH, CK, CM, JLW) acted as supervisors for the students although not all at the same time. Each instance of the course was supervised by one to three people.
The course was initiated with a brief summary of previously presented methods, motivations and objectives of the simulation.
In the first two hours the students were introduced to \python{} and \vpython{}.
In the following part the students worked with an initial code template (see~\cite{scripts}) and were motivated to come up with own solutions to fulfill the objectives.
A large part of the material covered during this day is presented in Sec.~\ref{sec:Numerical Implementation}.
With the exception of some younger students, the participants where able to simulate and visualize the motion of mercury including forces from GR.\@
Only a few participants started to implement the quantitative extraction of the perihelion motion, uncertainty estimates were not a subject as the time frame was too short.
Once the basic problem was tackled some students deviated from the suggested path and for instance studied the problem with one additional planet or different starting parameters.
This is a very interesting problem by itself for there are parameter ranges where the classical three-body system develops chaotic features.
We encouraged such explorations as well.
We include the additional material in the paper because we want to provide teachers with sufficient material to fill a working group outside the general curriculum running for a longer period than one day.

@@

\begin{figure}[htb]
\lstinputlisting[language=Python,basicstyle=\linespread{1}\footnotesize\ttfamily]{mercury_orbit_base.py}
\end{figure}
Further examples and template files can be found online~\cite{scripts}.
